Seuss Assembly
========================================================

Lets try an experiment where we try to assemble Green Eggs and Ham by Dr. Seuss from sentence fragments. To make it a bit harder, we will start by removing all punctuation (except Sam-I-am) and making everything upper case. Then load in the Dr. Seuss text into an array of words called `seuss`. I'm hiding this in the output, but you can see it in the source code for this document.

```{r seuss, include=FALSE}
seuss <- strsplit("IT IS INTERESTING TO CONTEMPLATE A TANGLED BANK CLOTHED WITH MANY PLANTS OF MANY KINDS WITH BIRDS SINGING ON THE BUSHES WITH VARIOUS INSECTS FLITTING ABOUT AND WITH WORMS CRAWLING THROUGH THE DAMP EARTH AND TO REFLECT THAT THESE ELABORATELY CONSTRUCTED FORMS SO DIFFERENT FROM EACH OTHER AND DEPENDENT UPON EACH OTHER IN SO COMPLEX A MANNER HAVE ALL BEEN PRODUCED BY LAWS ACTING AROUND US THESE LAWS TAKEN IN THE LARGEST SENSE BEING GROWTH WITH REPRODUCTION INHERITANCE WHICH IS ALMOST IMPLIED BY REPRODUCTION VARIABILITY FROM THE INDIRECT AND DIRECT ACTION OF THE CONDITIONS OF LIFE AND FROM USE AND DISUSE A RATIO OF INCREASE SO HIGH AS TO LEAD TO A STRUGGLE FOR LIFE AND AS A CONSEQUENCE TO NATURAL SELECTION ENTAILING DIVERGENCE OF CHARACTER AND THE EXTINCTION OF LESS IMPROVED FORMS THUS FROM THE WAR OF NATURE FROM FAMINE AND DEATH THE MOST EXALTED OBJECT WHICH WE ARE CAPABLE OF CONCEIVING NAMELY THE PRODUCTION OF THE HIGHER ANIMALS DIRECTLY FOLLOWS THERE IS GRANDEUR IN THIS VIEW OF LIFE WITH ITS SEVERAL POWERS HAVING BEEN ORIGINALLY BREATHED BY THE CREATOR INTO A FEW FORMS OR INTO ONE AND THAT WHILST THIS PLANET HAS GONE CIRCLING ON ACCORDING TO THE FIXED LAW OF GRAVITY FROM SO SIMPLE A BEGINNING ENDLESS FORMS MOST BEAUTIFUL AND MOST WONDERFUL HAVE BEEN AND ARE BEING EVOLVED", split=" ")[[1]]
#remove blanks
seuss <- seuss[seuss!=""]
```
Looking just at the start of the array:
```{r headSeuss}
head(seuss)
```
##Dot plot
Lets make a dot plot of Dr. Seuss. To do this we need a matrix of word positions and all the places that they are equal  to each other.
```{r dotmatrix}
library(reshape2)
library(ggplot2)
smatrix <- sapply(1:length(seuss), function(s){seuss==seuss[s]})
points <- melt(smatrix)
dots <- subset(points, value)
qplot(data = dots, 
      x = Var1, y = Var2, 
      xlab = "Word", ylab = "Word", 
      size = I(1), asp = 1)
```
That is a bit of a mess, so lets make it nicer and more informative by not using individual words, but phrases of some length. We'll write a function to do this, so we could do it for any text, and any phrase length. Then we'll plot the positions where phrase lengths of 3, 5, 10, and 25 words repeat.
```{r dotplot3}
phraseDotPlot <- function(words, l){
  phrases <- sapply(1:(length(words) - (l - 1)),
                 function(p){paste0(words[p:(p+l-1)], collapse = " ")})
  phrase_matrix <- sapply(1:length(phrases), function(p){phrases==phrases[p]})
  points <- melt(phrase_matrix)
  dots <- subset(points, value)
  qplot(data = dots, 
      x = Var1, y = Var2, 
      xlab = "Phrase", ylab = "Phrase", 
      size = I(1), asp = 1, main = sprintf("%d word phrases", l)
      )
}

phraseDotPlot(seuss, 3)
phraseDotPlot(seuss, 5)
phraseDotPlot(seuss, 10)
phraseDotPlot(seuss, 25)
```


Lets start with random 3 word phrases, but only from the first 133 words. So we can start with any position from 1:131, then take the next three words, concatenate them (with a space), and those will be our "reads". Lets start with 100 reads.

```{r threewords}
l = 3
n = 500
starts <- sample(1:(length(seuss)-(l-1)), size = n, replace = T)
fragments <- sapply(starts, function(s){paste0(seuss[s:(s+l-1)], collapse = " ")})

head(fragments)
```

