---
layout: Rmd
title: knitr markdown test
---
`r opts_chunk$set(tidy=FALSE, fig.path="images/Rplot-", fig.width=3, fig.height=3)`
This is a minimal example of using **knitr** with in HTML pages. I am actually
using markdown here since it is more convenient in GitHub.

First, the input file was named as `knitr-minimal.Rmd`
([source](https://github.com/yihui/knitr/blob/master/inst/examples/knitr-minimal.Rmd)),
and **knitr** will automatically determine the output filename to be
`knitr-minimal.md` (`*.Rmd --> *.md`).

I used the code below to make sure **knitr** writes correct URL's for my images.

``` {r setup}
opts_knit$set(base.dir="..", base.url="/")
render_jekyll()
hook_plot_md_side <- function(x, options){
  if (!is.null(options$fig.sidebar)) {
    paste(sub("\\s+$", "", hook_plot_md(x, options)),
          "{: .sidefig .align-right}\n",
          sep = ""
          )
  }else{
    hook_plot_md(x, options)
  } 
}
knit_hooks$set(plot = hook_plot_md_side)
````

Now we write some code chunks in this markdown file:

``` {r}
## a simple calculator
1+1
## boring random numbers
set.seed(123)
rnorm(5)
````

We can also produce plots:

``` {r md-cars-scatter, message=FALSE, fig.sidebar=T}
library(ggplot2)
qplot(hp, mpg, data=mtcars)+geom_smooth()
````

So no more hesitation on using GitHub and **knitr**! You just write a minimal
amount of code to get beautiful output on the web.